### 2014-05-16 ApTranscriptome_TR.Rmd
Cleaning up TR

**May 14, 2014**

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*.

@@

Once this is done, quantify expression for the Trimmomatic filtered reads from each colony-treatment sample separately. Note that for each sample, there are three four filtered read files:

@@

The TPM column for each sample was extracted and combined into a matrix for each colony.

@@

DeconSeq identified contaminant genes. I remove this n

@@

$$ log(TPM + 1) = \beta_0 + \beta_1(colony) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(colony * temp) + \beta_5(colony * temp^2) + \epsilon $$

@@

The set of transcripts with significant expression patterns include those with expression that differs by colony, temperature and the interaction of colony and temperature. In this section, I am specifically interested in the thermally-responsive transcripts (temperature and colony x temperature) so I subset the significant transcripts to examine these.

@@

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on colony, temperature or their interaction at 5% FDR.")
```

@@

I do this first for the thermally-responsive transcripts where there is no interaction with colony. For the transcripts where thermal-responsive expression depends on colony, I determine the functional type of the expression response separately for each colony.

@@

pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.", split.tables = "Inf")
```

Table 4 shows the number of transcripts that fall into each expression type for each each colony. The totals for each colony include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.

## Colony-level comparison ##

In this section, I compare the thermal reactionome between the *Ar* and *A22* colonies.

### Plasticity versus basal expression

Selection may have acted in response to thermal stress such that some genes are basaly activated in one colony while plastically-expressed in the other. I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony. Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*?

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at 19.25C in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

@@

Next I test the converse, that transcripts that are up-regulated at low temperatures in *Ar* are more highly-expressed at 19.25C temperatures in *A22*.

@@

Counter to expectations, expression at 19.5C is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.

@@

### Thermal tolerance indicated by region of constant gene expression ###

The 'Intermediate' expressed transcripts are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermall stress. We hypothesized that if the more southern *Ar* colony was more thermally-tolerant than *A22*, transcripts with 'Intermediate' expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each colony.

```{r thermal_breadth}
# extract 'Intermediate' expressed transcripts for A22 colony
A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
length(A22.int.lm)

@@

# T-test comparing standard deviation of expression between colonies
(t1 <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
```

Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

@@

### Thermal sensitivity indicated by response of bimodally-expressed transcripts ###

As the converse of the above hypothesis, a colony that is especially thermally-sensitive is likely to activate expression of molecular processes more quickly. We tested this using the same approach as for the 'Intermediate' transcripts, but using the inverse of the 'Bimodal' expressed transcripts.

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 colony
A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
length(A22.bim.lm)

@@

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
(t2 <- t.test(Ar.bim.sd, A22.bim.sd))
```

@@

### Compare peak expression among colonies ###

For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect? To test this, I use the framework of `r citet("10.1086/675302")`:

- *Offset, O*: overall difference in the mean expression value across all temperatures
- *Slope, S*: difference in overall slope
- *Curvature, C*: average difference in curvature of the reaction norm
- *Wiggle, W*: variability in shape not captured by the previous three measures

```{r offset, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)

# plot
boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))

t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))

# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
# offset = mean1 - mean2
varshape.out$offset <- varshape.out$A22.mean - varshape.out$Ar.mean

# t-test
(t.offset <- t.test(varshape.out$s.offset, alternative = "two.sided"))

# slope = slope1 - slope2
varshape.out$slope <- varshape.out$A22.slope - varshape.out$Ar.slope

# t-test
(t.slope <- t.test(varshape.out$s.slope, alternative = "two.sided"))

# curvature = curvature1 - curvature2
varshape.out$curvature <- varshape.out$A22.curve - varshape.out$Ar.curve
# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
varshape.out$s.curvature <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.curvature"] <- (varshape.out[i, "curvature"] / mean(c(varshape.out[i, "A22.curve"], varshape.out[i, "Ar.curve"])))
}
# change 'Inf' values to 0
varshape.out[which(varshape.out$Ar.curve == 0 & varshape.out$A22.curve == 0), "s.curvature"] <- 0
# t-test
(t.curvature <- t.test(varshape.out$s.curvature, alternative = "two.sided"))

# wiggle
varshape.out$wiggle <- varshape.out$A22.wiggle - varshape.out$Ar.wiggle
varshape.out$s.wiggle <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.wiggle"] <- (varshape.out[i, "wiggle"] / mean(c(varshape.out[i, "A22.wiggle"], varshape.out[i, "Ar.wiggle"])))
}
# change 'Inf' values to 0
varshape.out[which(varshape.out$Ar.wiggle == 0 & varshape.out$A22.wiggle == 0), "s.wiggle"] <- 0
# note small number of genes with 'wiggle'
length(which(varshape.out$s.wiggle != 0))

# t-test
(t.wiggle <- t.test(varshape.out$s.wiggle, alternative = "two.sided"))
```

Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means (offset), slope, curvature and wiggle.

```{r partition_reaction_norm_variation}
# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
varshape.out$s.offset <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.offset"] <- (varshape.out[i, "offset"] / mean(c(varshape.out[i, "A22.mean"], varshape.out[i, "Ar.mean"])))
}
# to compare among transcripts that differ in scale of expression, standardize by overall mean for each transcript
varshape.out$s.slope <- NA
for(i in 1:nrow(varshape.out)) {
  varshape.out[i, "s.slope"] <- (varshape.out[i, "slope"] / mean(c(varshape.out[i, "A22.slope"], varshape.out[i, "Ar.slope"])))
}
# for transcripts where both samples had slope=0, change s.slope from Inf to 0 as there is truly no difference in slope
varshape.out[which(varshape.out$Ar.slope == 0 & varshape.out$A22.slope == 0), "s.slope"] <- 0

                                        # weighted mean-standardized values of each measure
mean(abs((varshape.out$s.offset)))
mean(abs((varshape.out$s.slope)))
mean(abs((varshape.out$s.curvature)))
mean(abs((varshape.out$s.wiggle)))

# take absolute value of each value and sum to get total differences between reaction norms
varshape.out$s.total <- abs(varshape.out$s.offset) + abs(varshape.out$s.slope) + abs(varshape.out$s.curvature) + abs(varshape.out$s.wiggle)

(t6 <- t.test(varshape.out$s.total, alternative = "two.sided"))

# variation in reaction norms due to differences in mean
varshape.out$prop.mean <- abs(varshape.out$s.offset) / varshape.out$s.total
varshape.out$prop.slope <- abs(varshape.out$s.slope) / varshape.out$s.total
varshape.out$prop.curve <- abs(varshape.out$s.curvature) / varshape.out$s.total
varshape.out$prop.wiggle <- abs(varshape.out$s.wiggle) / varshape.out$s.total

# Mean proportion and 95% CI of total variation of each measure
mean(varshape.out$prop.mean)
quantile(varshape.out$prop.mean, probs = c(0.05, 0.5, 0.95))

mean(varshape.out$prop.slope)
quantile(varshape.out$prop.slope, probs = c(0.05, 0.5, 0.95))

mean(varshape.out$prop.curve)
quantile(varshape.out$prop.curve, probs = c(0.05, 0.5, 0.95))

mean(varshape.out$prop.wiggle)
qwiggle <- quantile(varshape.out$prop.wiggle, probs = c(0.05, 0.5, 0.95))
```

## Temperature of gene activation

@@

Calculate T~on~ for *High* genes in each colony.

@@

for(i in unique(Ar.high.TPM.dt.sub$Transcript)) {
    subdf <- Ar.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
}   
```

Determine if T~on~ is greater in *A22* or *Ar* for *High* genes.

```{r T_on_high_ttest}
t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
```

Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *AcNC* than *ApVT*.

@@

Genes with increased expression at *High* temperatures are turned on at, on average, 1°C higher in *ApVT* than *AcNC*.

@@

@@

@@

**2) High**

Genes with *High* expression in both colonies

Use `selectTranscript` function to select transcripts from 'Ap.response.type' for GSEA.

```{r gsea_high, results = 'hide'}
# Select transcripts
Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)

# Run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
#pandoc.table(Ap.high.gsea)
# output for Revigo
write.table(Ap.high.gsea[,"GO.ID"], file = "results/Ap_high_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
```

**3) Low**

Genes with *Low* expression in both colonies

```{r gsea_low, results = 'hide'}
# Select transcripts
Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ap.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.low) <- names(Ap.geneList)
Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
# check correct number of values set to 1
table(Ap.geneList.low)

# Run GSEA
Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
#pandoc.table(Ap.low.BP.ResTable)
write.table(Ap.low.gsea[,"GO.ID"], file = "results/Ap_low_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
```

**4) Bimodal**

Genes with *Bimodal* expression in both colonies

```{r gsea_bimodal, results = 'hide'}
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)

# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
#pandoc.table(Ap.bim.gsea)
write.table(Ap.bim.gsea[,"GO.ID"], file = "results/Ap_bim_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
```

**5) Intermediate**

Genes with *Intermediate* expression in both colonies

```{r gsea_intermediate, results = 'hide'}
Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "int" transcripts
Ap.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.int) <- names(Ap.geneList)
Ap.geneList.int[(which(names(Ap.geneList.int) %in% Ap.int))] <- 1
# check correct number of values set to 1
table(Ap.geneList.int)

# Run GSEA
Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
#pandoc.table(Ap.int.gsea)
write.table(Ap.int.gsea[,"GO.ID"], file = "results/Ap_int_gsea.txt", row.names = FALSE, sep = "\t", quote = FALSE)
```

```{r gsea_table, echo=FALSE, eval=TRUE, results='asis'}
## merge results into single table
Ap.high.gsea$Type <- "High"
Ap.low.gsea$Type <- "Low"
Ap.bim.gsea$Type <- "Bimodal"
Ap.int.gsea$Type <- "Intermediate"
# combine
Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
# reorder
Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea1)[7] <- "P"
# display
Ap.gsea1[ , 1:3]
```

Next, I perform GSEA for genes in each functional type in one colony but not the other (e.g. the set difference) to gain insight on differences between the colonies.

**A22 'High' genes not in Ar**

```{r gsea_high_A22, results = 'hide'}
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
# check correct number of values set to 1
table(A22.geneList.high)

# Run GSEA
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
#pandoc.table(A22.high.gsea, split.table = Inf)
```

**Ar 'High' genes not in A22**

```{r gsea_high_Ar, results = 'hide'}
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
# check correct number of values set to 1
table(Ar.geneList.high)

# Run GSEA
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
#pandoc.table(Ar.high.gsea, split.table = Inf)
```

**A22 'Low' genes not in Ar**

```{r gsea_low_A22, results = 'hide'}
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
# check correct number of values set to 1
table(A22.geneList.low)

# Run GSEA
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
#pandoc.table(A22.low.gsea, split.table = Inf)
```

**Ar 'Low' genes not in A22**

```{r gsea_low_Ar, results = 'hide'}
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
# check correct number of values set to 1
table(Ar.geneList.low)

# Run GSEA
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
#pandoc.table(Ar.low.gsea, split.table = Inf)
```

**A22 'Bimodal' genes not in Ar**

>>>>

**May 16, 2014**

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*. For historical reasons, I refer to these species as colonies at times throughout this technical report.

@@

Once this is done, quantify expression for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are three four filtered read files:

@@

The TPM column for each sample was extracted and combined into a matrix for each species.

@@

With gene expression quantified using all reads, I now remove the transcripts identified as contaminants by DeconSeq.

@@

$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$

@@

The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature. In this section, I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these.

@@

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts of 99,861 total with expression that depends on species, temperature or their interaction at 5% FDR.")
```

@@

I do this first for the thermally-responsive transcripts where there is no interaction with species. For the transcripts where thermal-responsive expression depends on species, I determine the functional type of the expression response separately for each species.

@@

pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.", split.tables = "Inf")
```

Table 4 shows the number of transcripts that fall into each expression type for each each species. The totals for each species include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.

## Species comparison ##

In this section, I perform a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:

- compare profiles of the temperature of maximum expression for thermally-responsive transcripts 
- compare basal expression at optimal temperature for thermally-responsive genes
- compare thermal stability of *Intermediate* genes
- compare thermal sensitivity of *Bimodal* genes
- compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
- evaluate the extent to which thermal reaction norms differ between species by mean and shape

### Compare temperature of maximum expression between species

### Compare basal expression at optimal temperature for thermally-responsive genes

Genes upregulated in response to thermal stress in one species may have greater basal levels of expression in the other species that experiences those stressful conditions more often. To test this hypothesis, we compared expression levels near the optimal temperature (19.5C) between the two species for genes in that are either in the 'High' or 'Low' expression group in the other species. Specifically, do genes upregulated at high temperatures in *A22* have greater basal expression at optimal temperatures in *Ar*?

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

@@

Next I test the converse: do genes up-regulated at low temperatures in *Ar* have greater basal expression near optimal temperatures in *A22*?

@@

Counter to expectations, expression at optimal temperatures is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.

@@

### Compare thermal stability of *Intermediate* genes

*Intermediate* genes are those that have expression that is shut-down as conditions become stressful, likely non-essential molecular processes. We hypothesized that if the more southern *Ar* species was more thermally-tolerant than *A22*, transcripts with *Intermediate* expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each species.

```{r thermal_breadth}
# extract 'Intermediate' expressed transcripts for A22
A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
length(A22.int.lm)

@@

# T-test comparing standard deviation of expression between colonies
(t.thermbreadth <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
```

Consistent with our hypothesis, *Intermediate* transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

@@

### Compare thermal sensitivity of *Bimodal* genes

As the converse of the above hypothesis, a species that is especially thermally-sensitive is likely to activate expression of molecular processes for thermal tolerance more quickly. We tested this using the same approach as for the *Intermediate* transcripts, but using the inverse of the *Bimodal* expressed transcripts.

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 species
A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
length(A22.bim.lm)

@@

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
(t.themsens <- t.test(Ar.bim.sd, A22.bim.sd))
```

@@

### Compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes

@@

Calculate T~on~ for *High* genes in each species.

@@

for(i in unique(Ar.high.TPM.dt.sub$Transcript)) {
    subdf <- Ar.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
}

# determine if T~on~ is greater in *A22* or *Ar* for *High* genes.

t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
```

Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.

@@

Genes with increased expression at *Low* temperatures are on average turned on 1°C higher in *ApVT* than *AcNC*.

@@

### Evaluate the extent to which differences in thermal reaction norms are due to mean or shape

For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect? To test this, I rougly follow `r citet("10.1086/675302")` by defining differences among reation norms for individual genes due to changes in the overall mean, slope, curvature and all higher-order shape differences (i.e. wiggle).

- *Mean, M*: overall difference in the mean expression value across all temperatures
- *Slope, S*: difference in overall slope
- *Curvature, C*: average difference in curvature of the reaction norm
- *Wiggle, W*: variability in shape not captured by the previous three measures

```{r varshape, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)

# delta mean
boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))

t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
t.mean

# delta slope
boxplot(log(varshape.out$A22.slope), log(varshape.out$Ar.slope))
t.slope <- t.test(log(varshape.out$A22.slope + 0.1), log(varshape.out$Ar.slope + 0.1), paired = TRUE)
t.slope

# delta curvature
boxplot(log(varshape.out$A22.curve), log(varshape.out$Ar.curve))
t.curvature <- t.test(log(varshape.out$A22.curve + 0.1), log(varshape.out$Ar.curve + 0.1), paired = TRUE)
t.curvature

# delta wiggle - small values so use un-transformed
boxplot(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
t.wiggle <- t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle, paired = TRUE)
t.wiggle
```

Reaction norms differ between species by mean and slope, but not by curvature or wiggle.

Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species. To do this, I first standardize each of these parameters so they are directly comparable.

```{r partition_reaction_norm_variation}
# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
# mean = mean1 - mean2
varshape.out$lmean <- log(varshape.out$A22.mean) - log(varshape.out$Ar.mean)

# slope = slope1 - slope2
varshape.out$lslope <- log(varshape.out$A22.slope + 0.1) - log(varshape.out$Ar.slope + 0.1)

# take absolute value of each value and sum to get total differences between reaction norms
varshape.out$ltotal <- abs(varshape.out$lmean) + abs(varshape.out$lslope)

# variation in reaction norms due to differences in mean
varshape.out$prop.lmean <- abs(varshape.out$lmean) / varshape.out$ltotal
varshape.out$prop.lslope <- abs(varshape.out$lslope) / varshape.out$ltotal

# Mean proportion and 95% CI of total variation of each measure
mean(varshape.out$prop.lmean)
(qlmean <- quantile(varshape.out$prop.lmean, probs = c(0.05, 0.5, 0.95)))

mean(varshape.out$prop.lslope)
(qlslope <- quantile(varshape.out$prop.lslope, probs = c(0.05, 0.5, 0.95)))
```

From this analysis, about 3/4 of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.

@@

Perform GSEA for all thermally-responsive transcripts.

@@

As the molecular processes involved in cold and hot temperature tolerance are different, I perform GSEA separately for each response type and compile results in a single table.

```{r gsea_by_type, results = 'hide'}
## Genes with *High* expression in both colonies
Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)
# run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)

## Genes with *Low* expression in both colonies
Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ap.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.low) <- names(Ap.geneList)
Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
# check correct number of values set to 1
table(Ap.geneList.low)
# Run GSEA
Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)

## Genes with *Bimodal* expression in both colonies
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)
# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)

## Genes with *Intermediate* expression in both colonies
Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "int" transcripts
Ap.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.int) <- names(Ap.geneList)
Ap.geneList.int[(which(names(Ap.geneList.int) %in% Ap.int))] <- 1
# check correct number of values set to 1
table(Ap.geneList.int)
# Run GSEA
Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)

## merge results into single table
Ap.high.gsea$Type <- "High"
Ap.low.gsea$Type <- "Low"
Ap.bim.gsea$Type <- "Bimodal"
Ap.int.gsea$Type <- "Intermediate"
# combine
Ap.gsea <- rbind(Ap.high.gsea, Ap.low.gsea, Ap.bim.gsea, Ap.int.gsea)
# reorder
Ap.gsea1 <- Ap.gsea[,c("Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea1)[7] <- "P"
# write to file
write.csv(Ap.setdiff.gsea, file = paste(resultsdir, "Ap_setdiff_GSEA_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```

```{r gsea_by_type_table, echo = FALSE, results = 'asis'}
pandoc.table(Ap.gsea1[, 1:3], caption = "Enriched GO terms for each thermal-response category")
#print(xtable(Ap.gsea1[ , 1:3], comment = getOption("xtable.comment", FALSE)))
```

Next, I perform GSEA for genes in each functional type in one species but not the other (e.g. the set difference) to gain insight on differences between the colonies.

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)

# *Ar 'High' genes not in A22**
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)

# A22 'Low' genes not in Ar**
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
# Ar 'Low' genes not in A22**
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)

# A22 'Bimodal' genes not in Ar**
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)

# Ar 'Bimodal' genes not in A22**
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)

# A22 'Intermediate' genes not in Ar**
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
table(A22.geneList.int)
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)

# Ar 'Intermediate' genes not in A22**
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)

# combine into single table
A22.high.gsea$Type <- "High"
A22.low.gsea$Type <- "Low"
A22.int.gsea$Type <- "Intermediate"
A22.bim.gsea$Type <- "Bimodal"
A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
A22.gsea$Species <- "ApVT"

Ar.high.gsea$Type <- "High"
Ar.low.gsea$Type <- "Low"
Ar.bim.gsea$Type <- "Bimodal"
Ar.int.gsea$Type <- "Intermediate"
Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
Ar.gsea$Species <- "AcNC"
